Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2019-12-19 analysis/paper/paper.Rmd
add figures, update code, work on PCA

However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks convey important information. In addition, obvious and reliable landmark points may be unavailable for some complex, rounded or globular shaped structures. In those cases, outline approaches are more suitable to study shape by focusing on the overall shape of a morphological structure. One of the outline approaches is semi-landmarks method, which assigns points along an outline curve at defined intervals between two landmarks to capture the variations of the outlines [@Bookstein1997; @Lawing2010]. Another approach, also the method used in this paper, is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adam2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometry of an object. The harmonic power expressed by the cumulated sum of squared harmonic coefficient usually provides a robust rule for choosing desired number of harmonics [@Bonhomme2014]. The application of EFA in archaeology is mostly used for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Outline approach based on EFA is suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variations, for example, ceramics. @Wilczek2014 evaluate the concordance between outline-based approaches, the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. They found that the variation in ceramic shapes analysed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes. In addition, the results of shape variation by EFA helps to understand the level of production standardisation over time across that region.

@@

phases_isect <- 
  pottery_data %>%  
  select(filename, phase) %>% 
  filter(filename %in% isect) %>% 
  mutate_all(as.factor) %>% 
  distinct(filename, phase) %>% 
  mutate(phase = as.factor(case_when(
    phase == 'post-e' ~ "Post-European",
    phase == 'pre-e' ~ "Pre-European",
    phase == 'ch-con' ~ "Post-European" # we combine Chinese and post-Euro
  )))

# how many pots in each phase?
phases_isect %>% 
  group_by(phase) %>% 
  tally(sort = T)

# start shape exploration
library(Momocs)
#  it freezes on the empty ones...
jpgs_imported <- import_jpg(jpgs, threshold = 0.5)

@@

# mean shape, per group
bot.ms <- mshapes(op, 2)
bot.ms$shp$`Post-European` <- bot.ms$shp$`Post-European`  %>% coo_plot(border="blue")
`Post-European`  <- bot.ms$shp$`Post-European`  %>% coo_draw(border="blue")
`Pre-European` <-   bot.ms$shp$`Pre-European`  %>% coo_draw(border="red")
legend("topright", 
       lwd=1,
       col=c("blue", "red"), 
       legend=c( "Post-European", "Pre-European"))

# or from mshapes directly
jpgs_imported_coo %>% efourier(6) %>% mshapes(~phase) %>% plot_mshapes()

# Thin plate spline analysis (TPS)
# Thin plate splines (TPS) was used to compare average shapes and visualise the outline deformations required to pass from an extreme of the morphospace to the other.
# Deformation grids
png(here::here("analysis/figures/tps-pre-Euro-to-post-Euro.png"), res = 300, w = 900, h = 900)
tps_grid(`Pre-European`, `Post-European` , shp.lwd = c(2, 2), shp.border = c("red", "blue"),
         amp = 1, grid.size = 18, legend = T)
dev.off()

png(here::here("analysis/figures/tps-pre-Euro-to-Chinese.png"), res = 300, w = 900, h = 900)
tps_grid(`Pre-European`, `Post-European`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
         amp = 1, grid.size = 18, legend = T)
dev.off()

@@

```{r PCA-analysis}
# pca
# these are fast
op.p <- PCA(op)

# We can see the pots groups by phase, and their shape distribution
# according to the first two principle components of the EFA 
# PC1 and PC2 explain 79% of variation
png("index/data/pottery-shape/pca-biplot.png", res = 300, w = 1500, h = 1500)
plot(op.p, 2, 
     morpho = TRUE, 
     ellipses = TRUE,
     cex = 1, 
     cex.labelsgroups = 1.5)

@@

# different visualization 
plot(PCA(op), 2, pos.shp="full")
plot(PCA(op), 2, pos.shp="range")
plot(PCA(op), 2, pos.shp="xy")
plot(PCA(op), 2, pos.shp="range_axes")
plot(PCA(op), 2, pos.shp="full_axes")

@@

ggsave("index/data/pottery-shape/pca-contrib.png")
scree_plot(op.p)

@@

@@

Outlines generated by EFA are examined by PCA, which shows the first three principal components (PCs) can explain 86% of shape variation. The first PC is related to the height of the vessels from short to high, the second PC reflects the width of their necks and mouths from narrow to wide, and the third PC relates to the degree of flare of neck from curve to straight. The biplot of PCs presents the grouping by each occupation phase. Multivariate analysis of variance (MANOVA) is used to test the difference in three shape varaibles from three phases. The result shows significant differences in shape between Pre-European and Post-European (p = 0.02) and Pre-European and Chinese contact shapes (p < 0.01). The distributions of the first three PCs for each vessel suggest variations in shape standardization by period. The first PC, the hight of the vessels, shows more variations in the pre-Euro period compared to the Chinese period. Standardization appears strong in the pre-Euro period but weak in the Chinese period. P-values for a modified signed-likelihood ratio test of equality of CVs of PC1 & PC2 show significant differences in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European.

>>>>

However, one limitation of landmark-based morphometrics is that landmark data may not be able to capture the shape differences of a morphological structure where the curving outlines between landmarks convey important information. In addition, obvious and reliable landmark points may be unavailable for some complex, rounded or globular shaped structures. In those cases, outline approaches are more suitable to study shape by focusing on the overall shape of a morphological structure. One of the outline approaches is semi-landmarks method, which assigns points along an outline curve at defined intervals between two landmarks to capture the variations of the outlines [@Bookstein1997; @Lawing2010]. Another approach is elliptic Fourier Analysis (EFA) commonly used for two-dimensional closed shape which turns coordinates into coefficients [@Kuhl1982]. EFA uses periodic functions to capture geometric information, where an outline is decomposed into a series of ellipses described by trigonometric functions, called harmonic coefficients [@Adam2004; @Claude2008; @Bonhomme2014]. The number of harmonics determines the quality and precision of the geometry of an object. The harmonic power, a cumulated sum of squared harmonic coefficient, provides a robust rule for desired number of harmonics [@Bonhomme2014]. The application of EFA in archaeology is mostly used for bone morphology to understand the biological variation and recently more for stone artefacts to explore stone tool technology [@Fox2015; @Hoggard2019; @Iovictua2010]. Outline approach based on EFA is suitable for shapes lacking obvious landmarks or curves convey more important information indicating the meaningful variations, for example, ceramics. @Wilczek2014 evaluate the concordance between outline-based approaches, the EFA and the Discrete Cosine Transform (DCT), and traditional typology by studying 154 complete ceramic vessels with varied shapes from the Bibracte oppidum in France. They found that the variation in ceramic shapes analysed by EFA and DCT matches traditional ceramic typology, which supports that outline-based approaches can be efficiently used for studying variations in ceramic shapes. In addition, the results of shape variation by EFA helps to understand the level of production standardisation over time across that region.

@@

phases_isect <- 
  pottery_data %>%  
  select(filename, phase) %>% 
  filter(filename %in% isect) %>% 
  mutate_all(as.factor) %>% 
  distinct(filename, phase) %>% 
  mutate(phase = as.factor(case_when(
    phase == 'post-e' ~ "Post-European",
    phase == 'pre-e' ~ "Pre-European",
    phase == 'ch-con' ~ "Chinese" # we combine Chinese and post-Euro
  )))

# how many pots in each phase?
phases_isect %>% 
  group_by(phase) %>% 
  tally(sort = T)
```

```{r jpg-coordinates, eval = FALSE}
# start shape exploration
library(Momocs)
#  it freezes on the empty ones...
jpgs_imported <- import_jpg(jpgs, threshold = 0.5)

@@

# mean shape, per group
pot.ms <- mshapes(op, 2)
pot.ms$shp$`Post-European` <- pot.ms$shp$`Post-European` %>% coo_plot(border="blue")
`Post-European` <- pot.ms$shp$`Post-European` %>% coo_draw(border="blue")
`Pre-European` <- pot.ms$shp$`Pre-European`  %>% coo_draw(border="red")
`Chinese` <- pot.ms$shp$`Chinese`  %>% coo_draw(border="green")
legend("topright", 
       lwd=1,
       col=c("red", "blue", "green"), 
       legend=c( "Pre-European", "Post-European", "Chinese"))

# from mshapes directly
jpgs_imported_coo %>% efourier() %>% mshapes(~phase) %>% plot_mshapes()

# Thin plate spline analysis (TPS)
# Thin plate splines (TPS) was used to compare average shapes and visualise the outline deformations required to pass from an extreme of the morphospace to the other.
# Deformation grids
png(here::here("analysis/figures/tps-pre-Euro-to-post-Euro.png"), res = 300, w = 900, h = 900)

tps_grid(`Pre-European`, `Post-European`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
         amp = 1, grid.size = 18, legend = T)
dev.off()

png(here::here("analysis/figures/tps-pre-Euro-to-Chinese.png"), res = 300, w = 900, h = 900)
tps_grid(`Pre-European`, `Chinese`, shp.lwd = c(2, 2), shp.border = c("red", "blue"),
         amp = 1, grid.size = 18, legend = T)
dev.off()

@@

```{r PCA-analysis}
# these are fast
op.p <- PCA(op)
# proportion of variances explained, PC1 and PC2 explain 79% of variation
pro_vari <-scree_plot(op.p)$data
vari_two <- round(pro_vari$cumsum[2],4)*100
vari_three <- round(pro_vari$cumsum[3],4)*100

# the pots groups by phase, and their shape distribution according to 
# the first two principle components of the EFA 
png(here::here("analysis/figures/pca-biplot.png"), res = 300, w = 1500, h = 1500)
plot(op.p, 2, 
     morpho = TRUE, 
     ellipses = TRUE,
     cex = 1, 
     cex.labelsgroups = 1.5)

@@

# different visualization 
plot(PCA(op), 2, pos.shp="full")
plot(PCA(op), 2, pos.shp="range")
plot(PCA(op), 2, pos.shp="xy")
plot(PCA(op), 2, pos.shp="range_axes")

plot(PCA(op), 2, pos.shp="full_axes")

@@

ggsave(here::here("analysis/figures/pca-contrib.png"))

@@

Outlines generated by EFA are examined by principal components analysis to explore the shape variations across three phases. The first two principal components (PCs) can explain `r vari_two`% of variances, and the first three principal components (PCs) can explain `r vari_three`% of variances. The first PC is related to the height of the vessels from short to high, the second PC reflects the width of their necks and mouths from narrow to wide, and the third PC relates to the degree of flare of neck from curve to straight. The biplot of PCs presents the grouping by each occupation phase.

@@

Multivariate analysis of variance (MANOVA) is used to test the difference in three shape varaibles from three phases. The result shows significant differences in shape between Pre-European and Post-European (p = 0.02) and Pre-European and Chinese contact shapes (p < 0.01). The distributions of the first three PCs for each vessel suggest variations in shape standardization by period. The first PC, the hight of the vessels, shows more variations in the pre-Euro period compared to the Chinese period. Standardization appears strong in the pre-Euro period but weak in the Chinese period. P-values for a modified signed-likelihood ratio test of equality of CVs of PC1 & PC2 show significant differences in standardization of vessel shape across some periods, especially between Chinese contact with either pre-European or post-European.
